---
title: "Data and Stuff"
author: "Duncan Gates"
navlink: "[More Data Stuff](http://duncangates.me/)"
og:
  type: "article"
  title: "opengraph title"
  url: "optional opengraph url"
  image: "optional opengraph image link"
footer:
  - content: '[Shiny Apps](http://shinyapps.duncangates.me/) • [Github](http://www.github.com/dungates/)<br/>'
  - content: 'Copyright © Duncan Gates, 2021'
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: DGThemes::skeleton
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, fig.width = 10, fig.height = 10)
library(tidyverse)
library(DGThemes)
library(rvest)
theme_set(theme_duncan())
```

## Web Scraping

From [this website](http://www.alltime-athletics.com/men.htm) we compile a dataframe of the track & field event and link to the events best times. I change around some of the naming for later purposes and delete events that don't particularly matter like the Javelin throw with old specs when the javelin was lighter. I also get read of all the extra links for non-legal marks since those are completely irrelevant.

```{r}
site <- "http://www.alltime-athletics.com/men.htm"
site_html <- rvest::read_html(site)

titles <- site_html %>%
  html_elements("a") %>%
  html_text()

(links <- site_html %>%
  html_elements("a") %>%
    html_attr("href"))

athletics_df <- bind_cols(event = titles, link = links) %>%
  slice(-c(1:2)) %>%
  mutate(link = paste0("http://www.alltime-athletics.com/", link)) %>%
  filter(event != "Old specs") %>%
  filter(!str_detect(link, "no")) %>%
  mutate(event = if_else(str_detect(event, "100 metres|200 metres"), str_replace_all(event, "metres", "meters"), event)) %>%
  mutate(event = str_replace_all(event, "10000 meters track walk", "10000 metres track walk"))
```

Then we can loop over this list to get a nested dataframe in which each column has the all time men's best performances in each area. The regex here is absolutely filthy and I at first attempted to write a regex that would encapsulate the separation in one go before I realized this could be broken down into parts making it much easier. First I separate out as much as I can using spaces followed by numbers, plus, minus, plus and minus, and a series of three capital letters. The athlete and city name strings were much tougher to deal with since they are variable lengths with different patterns of capitalization and numerics, but after separation it becomes as easy as separating when there is a digit preceded by a space, or a capital letter preceded by a space.

```{r}
link <- athletics_df$link[1]

elements <- read_html(link) %>%
  html_elements("pre") %>%
  magrittr::extract(1)
hundredm <- elements %>%
  html_text2() %>%
  str_split(pattern = "\\n") %>%
  unlist() %>%
  as_tibble() %>%
  slice(c(-1)) %>%
  mutate(value = str_squish(value)) %>%
  separate(value, 
           into = c("Rank", "Time", "wind_athlete", "Nation", "DoB", "race_rank_location", "Date"), 
           remove = T,
           sep = " (?=\\d)| (?=\\+)| (?=-)| (?=\\±)| (?=[A-Z]{3})") %>%# Regex translates to space followed by a number, 
                             # or space followed by a plus, 
                             # or space followed by a minus, 
                             # or space followed by three capital letters
  separate(wind_athlete,
           into = c("Wind", "Athlete"),
           remove = T,
           sep = "(?<=\\d)\\s") %>% # space preceded by a digit
  separate(race_rank_location,
           into = c("Race Rank", "Location"),
           remove = T,
           sep = "(?<=\\d)\\s|(?<=[A-Z])\\s") # space preceded by a digit, or space preceded by a capital letter

```

Let's make a function to do this, utilizing several conditionals in order to deal with the different kinds of data that will appear.

```{r}
read_event <- function(link, title) {
  # Use title to determine column names for later
  if (str_detect(title, "meters|110m hurdles")) { # 100 meter, 200 meter, and 110m hurdles
    columns <- c("Rank", "Time", "Wind", "Athlete First", "Athlete Last", "Nation", "DoB", "Event Rank", "Location", "Date")
  } else if (str_detect(title, "Mile|steeplechase|walk|400m hurdles|metres")) { # All distance running events, relays, 400m hurdles, and walking
    columns <- c("Rank", "Time", "Athlete First", "Athlete Last", "Nation", "DoB", "Event Rank", "Location", "Date")
  } else if (str_detect(title, "Long jump|Triple jump")) { # Long jump and Triple jump
    columns <- c("Rank", "Distance", "Wind", "Athlete First", "Athlete Last", "Nation", "DoB", "Event Rank", "Location", "Date")
  } else if (str_detect(title, "put|throw")) { # Shot put, discus throw, hammer throw, javelin throw
    columns <- c("Rank", "Distance", "Athlete First", "Athlete Last", "Nation", "DoB", "Event Rank", "Location", "Date")
  } else if (str_detect(title, "High jump|Pole vault")) { # High jump and pole vault
    columns <- c("Rank", "Height", "Wind", "Athlete First", "Athlete Last", "Nation", "DoB", "Event Rank", "Location", "Date")
  } else if (str_detect(title, "Decathlon")) { # Decathlon
    columns <- c("Rank", "Score", "Athlete First", "Athlete Last", "Nation", "DoB", "Event Rank", "Location", "Date")
  }
  # Read provided link and get all "pre" elements, we only want the first one in each case
  elements <- read_html(link) %>%
    html_elements("pre") %>%
    magrittr::extract(1)
  # Get the raw text and split new lines by \n, then make a dataframe and remove the first, empty, row, finally make sure it is whitespace 
  # separated and then separate
  event <- elements %>%
    html_text2() %>%
    str_split(pattern = "\\n") %>%
    unlist() %>%
    as_tibble() %>%
    slice(c(-1)) %>%
    mutate(value = str_squish(value)) %>%
    separate(value, into = columns, sep = " ")
}

read_event(link = athletics_df$link[1], title = athletics_df$event[1]) %>%
  view()
```


## Analysis

```{r}

```

